---
title: "Uncovering Racial and Demographic Trends in Chicago Police Stops: An Exploratory Analysis"
author: "Team 3"
output: 
    html_document:
      code_folding: hide
      number_sections: TRUE
      toc: yes
      toc_depth: 3
      toc_float: yes
    pdf_document:
      toc: yes
      toc_depth: '3'
date: "`r Sys.date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r basic_libraries, include=FALSE}
# Importing required libraries for the EDA.
library(ezids)
library(ggplot2)
```

## Introduction 

The Chicago Police Stops dataset provides valuable insights into the police stop practices in Chicago, with a particular focus on potential racial and demographic disparities. This dataset was collected as part of the Stanford Open Policing Project and contains information about police stops in the city. It includes various aspects of both the subjects and officers involved in the stops, such as race, age, gender, and outcomes like searches and arrests. The dataset serves as a foundation for examining how race, gender, and officer demographics influence the likelihood of being stopped, searched, or arrested during a police stop.

## Source of the Dataset

This dataset comes from the Stanford Open Policing Project. The variables in the dataset include:

Variable |  Definition  
  :-:    |  :-- 
location | The location where the stop occurred
subject_race |	The race of the person stopped
subject_age | The age of the person stopped 
subject_gender | The gender of the person stopped
officer_race | The race of the officer conducting the stop
officer_age |	The age of the officer 
search_made |	Whether a search was conducted during the stop 
arrest_made |	Whether an arrest was made during the stop 
violation |	The violation that prompted the stop

Source link: https://openpolicing.stanford.edu/data/

```{r}


```
