---
title: "Uncovering Racial and Demographic Trends in Chicago Police Stops: An Exploratory Analysis"
author: "Team 3"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r basic_libraries, include=FALSE}
# Importing required libraries for the EDA.
library(ezids)
library(ggplot2)
```

# Introduction 

The Chicago Police Stops dataset provides valuable insights into the police stop practices in Chicago, with a particular focus on potential racial and demographic disparities. This dataset was collected as part of the Stanford Open Policing Project and contains information about police stops in the city. It includes various aspects of both the subjects and officers involved in the stops, such as race, age, gender, and outcomes like searches and arrests. The dataset serves as a foundation for examining how race, gender, and officer demographics influence the likelihood of being stopped, searched, or arrested during a police stop.

## Source of the Dataset

This dataset comes from the Stanford Open Policing Project. The variables in the dataset include:

Variable |  Definition  
  :-:    |  :-- 
location | The location where the stop occurred
subject_race |	The race of the person stopped
subject_age | The age of the person stopped 
subject_gender | The gender of the person stopped
officer_race | The race of the officer conducting the stop
officer_age |	The age of the officer 
search_made |	Whether a search was conducted during the stop 
arrest_made |	Whether an arrest was made during the stop 
violation |	The violation that prompted the stop

Source link: https://openpolicing.stanford.edu/data/

## Glimplse of Dataset

```{r}
data=read.csv("Chicago Police Stops Dataset.csv")
head(data, 10)
```
## Structure and Summary of Dataset
```{r}
str(data)
```

```{r}
xkablesummary(data, title = "Summary of Dataset")
```

# Cleaning Dataset

## Null Values
```{r}
colSums(is.na(data))
```

```{r}
#removing missing values
cleaned_data<-na.omit(data)
```

## Post Removal of Null Values
Look into the structure of dataset.
```{r}
str(cleaned_data)
```
Look into summary of dataset.
```{r}
xkablesummary(cleaned_data, title = "Summary of Cleaned Dataset")
```
Look into null values of dataset.
```{r}
colSums(is.na(cleaned_data))
```
# Just An Initial Exploration
```{r}
barplot(table(cleaned_data$subject_race), main="Distribution of Stops by Race", col="lightblue")
```

# Detecting and Removing Outliers
```{r}
boxplot(cleaned_data$subject_age ~ cleaned_data$subject_race, main="Subject Age Distribution by Race")
```

```{r}
boxplot(cleaned_data$officer_age ~ cleaned_data$officer_race, main="Officer Age Distribution by Age")
```

```{r}
cleansubject_age=outlierKD2(cleaned_data,subject_age,rm=TRUE,qqplt = TRUE)
```

```{r}
cleanofficer_age=outlierKD2(cleaned_data,officer_age,rm=TRUE,qqplt = TRUE)
```

**'outlierKD2'** has identified and removed outliers, but it did not actually drop the rows from the dataframe. The outlierKD2 function only returns a modified column but does not update the dataframe in place. We manually remove the rows that contain outliers from cleaned_data dataframe.

```{r}
# Removing outliers from subject_age using IQR
Q1_subject <- quantile(cleaned_data$subject_age, 0.25)
Q3_subject <- quantile(cleaned_data$subject_age, 0.75)
IQR_subject <- Q3_subject - Q1_subject

# Define lower and upper bounds for subject_age
lower_bound_subject <- Q1_subject - 1.5 * IQR_subject
upper_bound_subject <- Q3_subject + 1.5 * IQR_subject

# Filter out rows with outliers in subject_age
cleaned_data <- cleaned_data[cleaned_data$subject_age >= lower_bound_subject & cleaned_data$subject_age <= upper_bound_subject, ]

# Removing outliers from officer_age using IQR
Q1_officer <- quantile(cleaned_data$officer_age, 0.25)
Q3_officer <- quantile(cleaned_data$officer_age, 0.75)
IQR_officer <- Q3_officer - Q1_officer

# Define lower and upper bounds for officer_age
lower_bound_officer <- Q1_officer - 1.5 * IQR_officer
upper_bound_officer <- Q3_officer + 1.5 * IQR_officer

# Filter out rows with outliers in officer_age
cleaned_data <- cleaned_data[cleaned_data$officer_age >= lower_bound_officer & cleaned_data$officer_age <= upper_bound_officer, ]

```

## Outliers Are Removed
```{r}
nrow(cleaned_data)
```
## Extreme Values Removed
```{r}
summary(cleaned_data$subject_age)
```

```{r}
summary(cleaned_data$officer_age)
```

```{r}
# Removing rows where subject_race or officer_race is "unknown"
cleaned_data <- cleaned_data[!(cleaned_data$subject_race == "unknown" | cleaned_data$officer_race == "unknown"), ]
```

# Hypothesis Testing 
```{r}
# Ensuring subject_race and officer_race are factors
cleaned_data$subject_race <- as.factor(cleaned_data$subject_race)
cleaned_data$officer_race <- as.factor(cleaned_data$officer_race)
```

*Disclaimer: Our dataset contained only cases where arrests were made (i.e., the arrest_made column only has "TRUE" values). The lack of variation (no FALSE values in arrest_made) made it hard to explore different hypotheses.*

However, we still did proceed with Chi-Square and ANOVA tests to explore relationships between subject race and officer race. 

## Chi - Square Test

*Null Hypothesis (H₀)*: There is no association between the subject's race and the officer's race during police stops where arrests were made.

*Alternative Hypothesis (H₁)*: There is a significant association between the subject's race and the officer's race during police stops where arrests were made.

```{r}
# Chi-Square test for association between subject_race and officer_race
chisq_result <- chisq.test(table(cleaned_data$subject_race, cleaned_data$officer_race))

print(chisq_result)
```

The Chi-Square test was performed to assess whether there is an association between subject race and officer race.

*Note: The warning sign typically occurs when the expected frequencies in some categories are very low. The test can still provide meaningful insights, but we keep limitation in mind.*

The *p-value* is *extremely small* (less than 0.05), hence we *reject* the *null hypothesis*.

There is a significant association between subject race and officer race. It implies that an officer's race may influence the decision to arrest subjects based on their race.

## ANOVA Test

*Null Hypothesis (H₀)*: There is no significant effect of subject race, officer race, or their interaction on the likelihood of arrests during police stops.

*Alternative Hypothesis (H₁)*: Subject race, officer race, or their interaction has a significant effect on the likelihood of arrests during police stops. 

```{r}
# Performing ANOVA to check if officer race impacts subject race in determining arrests
anova_result <- aov(officer_age ~ subject_race * officer_race, data = cleaned_data)

summary(anova_result)
```

Since the *p-values* for subject race, officer race, and their interaction are all *highly significant* (p < 2e-16), we *reject the null hypothesis*. 

Thus, we conclude that there is strong evidence to suggest that the race of the subject, the race of the officer, and the interaction between the two significantly impact the outcome of arrests during police stops.