---
title: "Uncovering Racial and Demographic Trends in Chicago Police Stops: An Exploratory Analysis"
author: "Team 3"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r basic_libraries, include=FALSE}
library(ezids)
library(ggplot2)
library(dplyr)
library(leaflet)
```

# Introduction 

The Chicago Police Stops dataset provides insights into the police stop practices in Chicago, focusing on location and potential racial and demographic disparities. This dataset was collected as part of the Stanford Open Policing Project and contains information about police stops in the city. Stanford Open Policing Project's goal is to create a repository detailing the interactions between police officers and the public. It includes various aspects of the subjects and officers involved in the stops, such as race, age, gender, and outcomes like searches and arrests. Also includes the type of violation and location where it was committed. The dataset serves as a foundation for examining how race, gender, violation, and officer demographics influence the likelihood of being stopped, searched, or arrested during a police stop.

## Source of the Dataset

This dataset comes from the Stanford Open Policing Project. The variables in the dataset include:

Variable |  Definition  
  :-:    |  :-- 
location | The location where the stop occurred
lat | The latitude of location
lng | The longitude of location
subject_race |	The race of the person stopped
subject_age | The age of the person stopped 
subject_gender | The gender of the person stopped
officer_race | The race of the officer conducting the stop
officer_age |	The age of the officer 
officer_gender | The gender of the officer stopping
search_made |	Whether a search was conducted during the stop 
arrest_made |	Whether an arrest was made during the stop 
violation |	The violation that prompted the stop

Source link: https://openpolicing.stanford.edu/data/

## SMART Questions

**1)** How do the subject's race and the officer's race influence the likelihood of arrest during police stops in Chicago, as observed in the dataset ?

**2)** How does the officer's gender influence the likelihood of arresting individuals of different races during police stops in Chicago, as measured by the association between officer gender and subject race in the arrest dataset?

## Glimplse of Dataset

```{r}
data=read.csv("il_chicago_2023_01_26.csv")
head(data, 5)
```

## Structure and Summary of Dataset

```{r}
str(data)
```

```{r}
xkablesummary(data, title = "Summary of Dataset")
```
# Cleaning Dataset
```{r}
data2 <- data[!(data$subject_race %in% c("unknown", "other") | data$officer_race %in% c("unknown", "other")), ]
```
 
We removed "unknown" and "other" races from subject_race and officer_race.
```{r}
nrow(data2)
```
## Null Values
```{r}
colSums(is.na(data2))
```
```{r}
#removing missing values
cleaned_data<-na.omit(data2)
```

## Post Removal of Null Values
After we removed the null values, we glimpse into the structure and summary of dataset.
```{r}
str(cleaned_data)
```
```{r}
xkablesummary(cleaned_data, title = "Summary of Cleaned Dataset")
```

Cross checking for null values of dataset.
```{r}
colSums(is.na(cleaned_data))
```

# Detecting & Removing Outliers
```{r}
ggplot(cleaned_data, aes(x = subject_race, y = subject_age)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs(title = "Subject Age Distribution by Race", x = "Subject Race", y = "Subject Age") +
  theme_minimal()

```

```{r}
ggplot(cleaned_data, aes(x = officer_race, y = officer_age)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs(title = "Subject Age Distribution by Race", x = "Officer Race", y = "Officer Age") +
  theme_minimal()
```

```{r}
cleansubject_age=outlierKD2(cleaned_data,subject_age,rm=TRUE,qqplt = TRUE)
```

```{r}
cleanofficer_age=outlierKD2(cleaned_data,officer_age,rm=TRUE,qqplt = TRUE)
```

**'outlierKD2'** has identified and removed outliers, but it did not actually drop the rows from the dataframe. The outlierKD2 function only returns a modified column but does not update the dataframe in place. We manually remove the rows that contain outliers from cleaned_data dataframe.

```{r}

Q1_subject <- quantile(cleaned_data$subject_age, 0.25)
Q3_subject <- quantile(cleaned_data$subject_age, 0.75)
IQR_subject <- Q3_subject - Q1_subject


lower_bound_subject <- Q1_subject - 1.5 * IQR_subject
upper_bound_subject <- Q3_subject + 1.5 * IQR_subject


cleaned_data <- cleaned_data[cleaned_data$subject_age >= lower_bound_subject & cleaned_data$subject_age <= upper_bound_subject, ]
```

```{r}
Q1_officer <- quantile(cleaned_data$officer_age, 0.25)
Q3_officer <- quantile(cleaned_data$officer_age, 0.75)
IQR_officer <- Q3_officer - Q1_officer

lower_bound_officer <- Q1_officer - 1.5 * IQR_officer
upper_bound_officer <- Q3_officer + 1.5 * IQR_officer

cleaned_data <- cleaned_data[cleaned_data$officer_age >= lower_bound_officer & cleaned_data$officer_age <= upper_bound_officer, ]
```

## Outliers Are Removed
```{r}
nrow(cleaned_data)
```

## Extreme Values Removed
```{r}
summary(cleaned_data$subject_age)
```

```{r}
summary(cleaned_data$officer_age)
```

```{r}
top_violations <- cleaned_data %>%
  count(violation, sort = TRUE) %>%    
  top_n(10, n)                        


final_data <- cleaned_data %>%
  filter(violation %in% top_violations$violation)
```

We filtered the dataset to retain records of top 10 violations.

```{r}
str(final_data)
```

```{r}
head(final_data, 10)
```

```{r}
print(unique(final_data$violation))
```

# Visualization

## Bar Plot
```{r}
ggplot(final_data, aes(x = subject_race)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Distribution of Police Stops by Subject Race", x = "Subject Race", y = "Count") +
  theme_minimal()
```

The above bar plot shows the distribution of police stops by subject race, with African/American subjects having the highest count of arrests, followed by Hispanic/Latino and Caucasian subjects. The number of stops for Asian/Pacific Islander is significantly lower in comparison.

```{r}
ggplot(final_data, aes(x = officer_race)) +
  geom_bar(fill = "coral", color = "black") +
  labs(title = "Distribution of Police Stops by Officer Race", x = "Officer Race", y = "Count") +
  theme_minimal()
```

The above bar plot illustrates the distribution of police stops by officer race. It shows that Caucasian officers are involved in the highest number of police stops, followed by Hispanic/Latino and African American officers. Asian/Pacific Islander officers are involved in a relatively low number of stops.

## Violin Plot
```{r}
ggplot(final_data, aes(x = subject_race, y = subject_age, fill = subject_race)) +
  geom_violin(scale = "width", trim = FALSE) +
  geom_boxplot(width = 0.1, position = position_dodge(width = 0.9)) +
  labs(title = "Distribution of Subject Age by Race",
       x = "Subject Race", y = "Subject Age", fill ="Subject_Race") +
  theme_minimal()
```

This violin plot showcases distribution of subject age across different racial groups. The wider parts of the violin indicate higher densities of points; the narrow part indicates low-density concentrations for points at a certain age. For the four groups represented, the median age is mainly between 30 and 40 years. Asian/Pacific Islander subjects have the widest age spread, with some older outliers, while the Hispanic/Latino subjects have more concentrated and younger age range. 

## Box Plot
```{r}
ggplot(final_data, aes(x = officer_race, y = officer_age, fill = officer_race)) +
  geom_boxplot() +
  labs(title = "Distribution by Officer Age by Race", x = "Officer Race", y = "Officer Age") +
  theme_minimal()
```

The box plot shows distribution of officer age across different racial groups. The median age for all groups is somewhere in the early and mid-40s; African American officers are the exception, with a little bit higher median age than others. Asian/Pacific Islander and Hispanic/Latino officers demonstrate narrower distributions in terms of age, with most officers being close to age 40.  

## Stacked Bar Plot
```{r}
ggplot(final_data, aes(x = officer_age, fill = subject_race)) +
  geom_bar(position = "fill") +
  facet_wrap(~ officer_sex) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Subject Race by Officer Age and Gender",
       x = "Officer Age", y = "Proportion of Subjects Arrested",
       fill = "Subject Race") +
  theme_minimal()
```

The above stacked bar plot shows the proportion of different racial groups arrested by the age and gender of officers. In general, for both female and male officers, younger officers arrested more African-Americans and Hispanics/Latinos while older ones arrested an increased proportion of Caucasians and Asian/Pacific Islanders. Overall, this means that the racial distribution of arrests shifts with the age of the officer and varies slightly by gender.

## Heatmap
```{r}
race_correlation <- as.data.frame(table(final_data$subject_race, final_data$officer_race))
```

```{r}
ggplot(race_correlation, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "black") +  
  scale_fill_gradient(low = "white", high = "lightblue") +  
  labs(title = "Correlation between Subject and Officer Race", 
       x = "Subject Race", 
       y = "Officer Race", 
       fill = "Count") +
  theme_minimal()
```

The heatmap reveals the correlation between officer race and the race of subjects they interacted with. Darker blue colors signify greater frequency of occurrences. The Caucasian officers significantly interacted with African American subjects. Interaction with subjects from among the other officer races shows weaker correlations, evidenced by the lighter blue coloration of the cells in the matrix. Hispanic/Latino officers tend to stop more Hispanic/Latino subjects, while other race combinations occur less frequently, indicating potential racial correlations in police stops.

## Maps
```{r}
violation_pal <- colorFactor(palette = "Set3", domain = final_data$violation)

leaflet(data = final_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~lng, lat = ~lat, 
    color = ~violation_pal(violation),  
    popup = ~paste("Violation:", violation, "<br>",
                   "Subject Age:", subject_age, "<br>",
                   "Officer Age:", officer_age),
    radius = 5,
    stroke = FALSE,
    fillOpacity = 0.8
  ) %>%
  addLegend(
    position = "bottomright", 
    pal = violation_pal, 
    values = ~violation,
    title = "Violation Type"
  )
```
The map above gives us a good view of the top 10 violations across the city.

## Bar Plot
```{r}
top_race_per_violation <- final_data %>%
  group_by(violation, subject_race) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(desc(count)) %>%
  slice_max(order_by = count, by = violation, n = 1)

ggplot(top_race_per_violation, aes(x = violation, y = count, fill = violation)) +
  geom_bar(stat = "identity") +
  labs(title = "Top Subject Race by Violation",
       x = "Violation Type",
       y = "Count") +
  theme_minimal() +
  geom_text(aes(label = subject_race), angle = 90, vjust = -0.5, hjust = 0.5) +  
  theme(axis.text.x = element_blank(),  
        axis.title.x = element_blank()) +  
  scale_fill_brewer(palette = "Set3")
```
The above bar plot shows the subject race violating specific rules highest for each type of violation.